A Constraint Solving Approach to Parikh Images of Regular Languages

### 2022-07-08 main.tex
updated notation
Single example

\begin{figure}
  \centering
\begin{prooftree}
  \hypo{\bigstar}
  \infer1[]{
    \begin{matrix}
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 = 0 \land \\
      2 + \TransitionVar_4 > 0 \land \\
      k = 1 + \TransitionVar_4 
    \end{matrix}
  }
  \infer1[\Subsume{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 = 0 \land \\
      2 + \TransitionVar_4 > 0 \land \\
      k = 1 + \TransitionVar_4 
    \end{matrix}
  }
  \infer1[\Propagate]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 \geq 0 \land \\
      2 + \TransitionVar_4 \geq 0 \land \\
      2k = 2 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\EquationReasoning{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6
    \end{matrix}
  }
  % BRANCH:  = 0
  \hypo{\bigstar}
  \infer1[]{
    \begin{matrix}
      \TransitionVar_3 > 0 \\
      \TransitionVar_6 = 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\Subsume{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_3 > = 0 \\
      \TransitionVar_6 = 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\EquationReasoning{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_6 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0\\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\Propagate{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_6 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\EquationReasoning{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6
    \end{matrix} % BRANCH: > 0
  } % SPLIT
  \infer2[\Split{} $1 - \TransitionVar_6$]{ \Connected(\AcaOrBc{}, \Filter) \land 2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6 }
  \infer1[\EquationReasoning{}]{
    \Connected(\AcaOrBc{}, \Filter) \land
    2k + 1 =
    \sum\limits_{\Transition \in \Transitions} (\Transition \mapsto 1)(\Label) \cdot \Filter(\Transition)
  }
  \infer1[\Expand{}]{\Image{}_{\AcaOrBc{}, \Transition \mapsto 1}(\Filter, 2k + 1)}
\end{prooftree}
\caption{A derivation for \Calculus{} computing odd lengths in $\AcaOrBc{}$.}\label{fig:derivation:single}
\end{figure}

In \cref{fig:derivation:single}, we see how we start by expanding the predicate
using the (simplified) flow equations. We can also see how interleaving
reasoning on the corresponding linear equations helps the production. In both
branches the reasoning is similar: we conclude that when choosing either path at
the starting state we must avoid the other one, perform propagation based on
that fact, use some algebraic reasoning to derive the expected forms of the
upper and lower bounds to propagate the now disconnected transitions, and
finally subsume, removing the $\Connected{}$ predicate when we are unable to use
either of the rules, leaving only the satisfiable remnants of the flow equations
at the respective leaves marked with~$\bigstar$. To obtain a solution from the
proof, one would need to perform standard model generation on the remaining
constraints to obtain values for the sought variable $\MonoidElement$. The full
Parikh image is the disjunction of the constraints left at the two
$\bigstar$-marked leaves, after quantifier elimination.

@@

      \Image{}_{\AcaOrBc{}\times\SomethingCSomething{}, \Map}(\Filter, 

@@

  \infer1[\ExpandM]{\Image{}_{\AcaOrBc{}\times\SomethingCSomething{}, \Map}(\Filter, \VectorLiteral{a, b, c}) \land a > b}

>>>>

\begin{figure}
  \centering
\begin{prooftree}
  \hypo[]{
    \begin{matrix}
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 = 0 \land \\
      2 + \TransitionVar_4 > 0 \land \\
      k = 1 + \TransitionVar_4 
    \end{matrix}
  }
  \infer1[\Subsume{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 = 0 \land \\
      2 + \TransitionVar_4 > 0 \land \\
      k = 1 + \TransitionVar_4 
    \end{matrix}
  }
  \infer1[\Propagate]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      \TransitionVar_6 = 1 \land \\
      \TransitionVar_4 + 1 > 0 \land \\
      \TransitionVar_3 \geq 0 \land \\
      2 + \TransitionVar_4 \geq 0 \land \\
      2k = 2 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\EquationReasoning{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 = 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6
    \end{matrix}
  }
  % BRANCH:  = 0'
  \hypo{
    \begin{matrix}
      \TransitionVar_3 > 0 \land \\
      \TransitionVar_6 = 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\Subsume{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_3 > = 0 \land \\
      \TransitionVar_6 = 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\EquationReasoning{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_6 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2\TransitionVar_6 + \TransitionVar_4 = 0\\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\Propagate{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      \TransitionVar_6 = 0 \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3
    \end{matrix}
  }
  \infer1[\EquationReasoning{}]{
    \begin{matrix}
      \Connected(\AcaOrBc{}, \Filter) \land \\
      1 - \TransitionVar_6 > 0 \land \\
      2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6
    \end{matrix} % BRANCH: > 0
  } % SPLIT
  \infer2[\Split{} $1 - \TransitionVar_6$]{ \Connected(\AcaOrBc{}, \Filter) \land 2k = 1 + 2\TransitionVar_4 + \TransitionVar_3 + \TransitionVar_6 }
  \infer1[\EquationReasoning{}]{
    \Connected(\AcaOrBc{}, \Filter) \land
    2k + 1 =
    \sum\limits_{\Transition \in \Transitions} (\Transition \mapsto 1)(\Label) \cdot \Filter(\Transition)
  }
  \infer1[\Expand{}]{\Image{}_{\AcaOrBc{}, \Transition \mapsto 1}(\Filter, 2k + 1)}
\end{prooftree}
\caption{A derivation for \Calculus{} computing odd lengths in $\AcaOrBc{}$.}\label{fig:derivation:single}
\end{figure}

In \cref{fig:derivation:single}, we see how we start by expanding the predicate
using the (simplified) flow equations. We can also see how interleaving
reasoning on the corresponding linear equations helps the deduction. In both
branches the reasoning is similar: we conclude that when choosing either path at
the starting state we must avoid the other one, perform propagation based on
that fact, use some algebraic reasoning to derive the expected forms of the
upper and lower bounds to propagate the now disconnected transitions, and
finally subsume, removing the $\Connected{}$ predicate when we are unable to use
either of the rules, leaving only a set of linear equations. To obtain a solution from the
proof, one would need to perform standard model generation on the remaining
constraints to obtain values for the sought variable $\MonoidElement$. The full
Parikh image is the disjunction of the constraints left at the two
leaves after quantifier elimination.

Following the values of $\Filter$, we can read out the path corresponding to our solution in the product, and with that a word. In our case, for the right branch we have $\Filter(\FromLabelTo{S}{b}{B}) = 0,
\Filter(\FromLabelTo{A}{c}{A})  > 0,
\Filter(\FromLabelTo{B}{b}{S}) = 0,
\Filter(\FromLabelTo{S}{a}{A}) = 1 ,
\Filter(\FromLabelTo{A}{a}{F}) = 1 ,
\Filter(\FromLabelTo{B}{c}{F}) = 0$. We can read out a word along it: aca, accca, etc, and the corresponding constraint: $\exists k : \TransitionVar_3 = 2k -1 \land \TransitionVar_3 > 0 \land k \geq 0$.

@@

      \Image{}_{\Tuple{\AcaOrBc{},\SomethingCSomething{}}, \Map}(\Filter, 

@@

  \infer1[\ExpandM]{\Image{}_{\Tuple{\AcaOrBc{},\SomethingCSomething{}}, \Map}(\Filter, \VectorLiteral{a, b, c}) \land a > b}
